---
title: "Ordinal Logistic Regression: Diabetes Health Status"
subtitle: "Analysis of CDC BRFSS 2015 Health Indicators Dataset"
author: "DrPH Student"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: flatly
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This assignment applies **ordinal logistic regression** (proportional odds model) to the CDC Behavioral Risk Factor Surveillance System (BRFSS) 2015 dataset, which contains 253,680 survey responses from US adults.

**Research question:** What sociodemographic, behavioural, and clinical factors are associated with severity of diabetes status (No Diabetes → Pre-diabetes → Diabetes)?

**Outcome variable:** `Diabetes_012` : an ordered three-category variable:

- `0` = No diabetes
- `1` = Pre-diabetes
- `2` = Diabetes

The outcome has a natural ordering. That ordering is what justifies the **proportional odds model (cumulative link model)** here, following Hosmer, Lemeshow & Sturdivant (Chapter 8, *Applied Logistic Regression*, 3rd ed.).

---

# Data Preparation

## Load Libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(ordinal)     # clm() : cumulative link model
library(gtsummary)   # publication-quality tables
library(broom)       # tidy model outputs
library(knitr)       # table rendering
library(kableExtra)  # table styling
library(car)         # vif() : multicollinearity check
```

## Read and Inspect Data

```{r}
#| label: read-data

dat <- read_csv("diabetes_sample_2200.csv")
glimpse(dat)
```

The working dataset is a **stratified random sample** (n = 2,200) from the full CDC BRFSS 2015 dataset (n = 253,680). Stratification was on `Diabetes_012`:

- No Diabetes: n = 1,600 (72.7%)
- Pre-Diabetes: n = 150, oversampled (population prevalence ≈ 1.8%)
- Diabetes: n = 450 (20.5%)

Pre-diabetes was deliberately oversampled. Proportional sampling would yield fewer than 40 cases, too few for stable coefficient estimation. ORs here reflect sample-level associations and cannot be used to derive population prevalence.

## Variable Recoding

```{r}
#| label: recode-variables

dat <- dat %>%
  mutate(
    diabetes_cat = factor(Diabetes_012,
                          levels  = c(0, 1, 2),
                          labels  = c("No Diabetes", "Pre-Diabetes", "Diabetes"),
                          ordered = TRUE),

    HighBP               = factor(HighBP,               labels = c("No", "Yes")),
    HighChol             = factor(HighChol,             labels = c("No", "Yes")),
    CholCheck            = factor(CholCheck,            labels = c("No", "Yes")),
    Smoker               = factor(Smoker,               labels = c("No", "Yes")),
    Stroke               = factor(Stroke,               labels = c("No", "Yes")),
    HeartDiseaseorAttack = factor(HeartDiseaseorAttack, labels = c("No", "Yes")),
    PhysActivity         = factor(PhysActivity,         labels = c("No", "Yes")),
    Fruits               = factor(Fruits,               labels = c("No", "Yes")),
    Veggies              = factor(Veggies,              labels = c("No", "Yes")),
    HvyAlcoholConsump    = factor(HvyAlcoholConsump,    labels = c("No", "Yes")),
    AnyHealthcare        = factor(AnyHealthcare,        labels = c("No", "Yes")),
    NoDocbcCost          = factor(NoDocbcCost,          labels = c("No", "Yes")),
    DiffWalk             = factor(DiffWalk,             labels = c("No", "Yes")),
    Sex                  = factor(Sex,                  labels = c("Female", "Male")),

    # Unordered factor : produces interpretable category comparisons (ref = Excellent)
    GenHlth = factor(GenHlth,
                     levels = 1:5,
                     labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")),

    # Collapsed from 13 two-to-five-year groups into 6 clinical bands
    Age = factor(dplyr::case_match(Age,
                   c(1, 2, 3)    ~ "18-34",
                   c(4, 5)       ~ "35-44",
                   c(6, 7)       ~ "45-54",
                   c(8, 9)       ~ "55-64",
                   c(10, 11)     ~ "65-74",
                   c(12, 13)     ~ "75+"),
                 levels = c("18-34", "35-44", "45-54", "55-64", "65-74", "75+")),

    Education = factor(Education,
                       levels = 1:6,
                       labels = c("Never attended", "Elementary", "Some high school",
                                  "High school grad", "Some college", "College grad")),
    Income    = factor(Income,
                       levels = 1:8,
                       labels = c("<$10k", "$10-15k", "$15-20k", "$20-25k",
                                  "$25-35k", "$35-50k", "$50-75k", ">$75k"))
  )

levels(dat$diabetes_cat)
```

Ordering confirmed: **No Diabetes < Pre-Diabetes < Diabetes**.

---

# Exploratory Data Analysis

## Outcome Distribution

```{r}
#| label: outcome-distribution
#| fig-cap: "Distribution of diabetes status"

dat %>%
  count(diabetes_cat) %>%
  mutate(pct   = round(n / sum(n) * 100, 1),
         label = paste0(n, " (", pct, "%)")) %>%
  ggplot(aes(x = diabetes_cat, y = n, fill = diabetes_cat)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = c("#2ecc71", "#f39c12", "#e74c3c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  labs(title = "Distribution of Diabetes Status", x = NULL, y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))
```

## Descriptive Statistics by Outcome

```{r}
#| label: descriptive-table

dat %>%
  select(diabetes_cat, HighBP, HighChol, BMI, Smoker,
         PhysActivity, GenHlth, Age, Sex, Education, Income) %>%
  tbl_summary(
    by = diabetes_cat,
    statistic = list(all_continuous()  ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    label = list(
      HighBP       ~ "High Blood Pressure",
      HighChol     ~ "High Cholesterol",
      BMI          ~ "BMI (continuous)",
      Smoker       ~ "Ever Smoked ≥100 cigarettes",
      PhysActivity ~ "Physical Activity (past 30 days)",
      GenHlth      ~ "General Health (self-rated)",
      Age          ~ "Age Group",
      Sex          ~ "Sex",
      Education    ~ "Education Level",
      Income       ~ "Income Category"
    )
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  modify_caption("**Table 1. Descriptive Statistics by Diabetes Status**")
```

---

# Model Building

## Univariable Screening

All candidate predictors are screened individually using separate proportional odds models (`clm`). Variables with **p < 0.25** (Hosmer-Lemeshow criterion) are considered for the multivariable model.

```{r}
#| label: univariable-screening

dat %>%
  select(diabetes_cat,
         HighBP, HighChol, BMI, Smoker, PhysActivity, GenHlth, Age, Sex,
         Education, Income, CholCheck, Stroke, HeartDiseaseorAttack,
         Fruits, Veggies, HvyAlcoholConsump, AnyHealthcare, NoDocbcCost,
         DiffWalk) %>%
  tbl_uvregression(
    method       = clm,
    y            = diabetes_cat,
    exponentiate = TRUE,
    hide_n       = TRUE
  ) %>%
  bold_p(t = 0.25) %>%
  bold_labels() %>%
  modify_caption("**Table 2. Univariable Screening : Unadjusted Odds Ratios (bold = p < 0.25)**")
```

**Screening decision:** Bold p-values mark variables meeting the p < 0.25 entry criterion. From those, the multivariable model retains predictors selected on clinical relevance and established BRFSS literature: **High Blood Pressure, High Cholesterol, BMI, Smoking, Physical Activity, General Health, Sex, and Age group.** Variables passing the screen but not retained (e.g., Fruits, Veggies, HvyAlcoholConsump) did not meet the clinical relevance threshold for the primary model.

## Full Adjusted Model

The proportional odds model is fitted using `ordinal::clm()` with the logit link. Two sets of parameters are estimated simultaneously:

- **Threshold parameters** (α₁, α₂) : the two log-odds boundaries separating the three ordered categories (*No Diabetes|Pre-Diabetes* and *Pre-Diabetes|Diabetes*). These are nuisance parameters that define the model's internal cut-points and carry no direct clinical meaning.
- **Regression coefficients** (β) : one per predictor, shared across both thresholds under the proportional odds constraint. Positive β indicates association with more severe diabetes, and negative β is protective.

The `summary()` output has two blocks: (1) **Threshold coefficients** (the α cut-points) and (2) **Coefficients** (the β estimates with SE, z-statistic, and p-value). Clinical inference comes from the Coefficients block only.

```{r}
#| label: full-model

m_full <- clm(diabetes_cat ~ HighBP + HighChol + BMI + Smoker +
                PhysActivity + GenHlth + Sex + Age,
              data = dat,
              link = "logit")

summary(m_full)
```

**Reading the output:** Look at the **Coefficients** block. Positive coefficients (e.g., HighBP, BMI) indicate association with higher diabetes severity, and negative ones (e.g., PhysActivity) are protective. The threshold parameters (α) above that block define internal cut-points and are not tested.

---

# Model Diagnostics

## Multicollinearity Check

`ordinal::clm()` does not have a native VIF method. A proxy linear model fitted to the same predictors provides a well-accepted VIF approximation via `car::vif()`.

```{r}
#| label: vif-check

mod_lm_vif <- lm(as.numeric(diabetes_cat) ~ HighBP + HighChol + BMI + Smoker +
                   PhysActivity + GenHlth + Sex + Age,
                 data = dat)

vif_mat     <- car::vif(mod_lm_vif)
vif_adj     <- vif_mat[, 3]
vif_max     <- round(max(vif_adj), 2)
vif_top     <- names(which.max(vif_adj))
vif_concern <- vif_adj[vif_adj >= 2.0]
vif_mat
```

**Interpretation:** `r if(length(vif_concern) == 0) { paste0("All GVIF^(1/(2·Df)) values fall below 2.0. The highest adjusted VIF is ", vif_max, " for ", vif_top, ". No predictor shows problematic collinearity, thus coefficients and standard errors are interpreted without reservation.") } else { paste0(length(vif_concern), " predictor(s) exceed GVIF^(1/(2·Df)) = 2.0: ", paste(names(vif_concern), round(vif_concern, 2), sep = " = ", collapse = ", "), ". Standard errors for these predictors may be inflated, thus OR estimates are interpreted with caution. The highest adjusted VIF is ", vif_max, " for ", vif_top, ".") }`

## BMI Linearity Check

The proportional odds model assumes a log-linear relationship between BMI and diabetes severity on the log-odds scale. This is checked by computing the Pearson correlation between BMI quartile midpoints and empirical log-odds at both cumulative thresholds.

```{r}
#| label: bmi-linearity
#| fig-cap: "Empirical log-odds of P(Y ≥ j) across BMI quartiles."

bmi_lo <- dat %>%
  mutate(BMI_q = cut(BMI,
                     breaks = quantile(BMI, probs = seq(0, 1, 0.25), na.rm = TRUE),
                     include.lowest = TRUE)) %>%
  group_by(BMI_q) %>%
  summarise(
    BMI_mid = mean(BMI),
    lo_pre  = qlogis(pmin(pmax(mean(as.integer(diabetes_cat) >= 2), 1e-6), 1 - 1e-6)),
    lo_dia  = qlogis(pmin(pmax(mean(as.integer(diabetes_cat) >= 3), 1e-6), 1 - 1e-6)),
    .groups = "drop"
  )

bmi_cor_pre <- round(cor(bmi_lo$BMI_mid, bmi_lo$lo_pre), 3)
bmi_cor_dia <- round(cor(bmi_lo$BMI_mid, bmi_lo$lo_dia), 3)

tibble(Threshold   = c("≥ Pre-Diabetes", "≥ Diabetes"),
       Correlation = c(bmi_cor_pre, bmi_cor_dia)) %>%
  kable(caption = "Pearson correlation : BMI quartile midpoint vs empirical log-odds",
        digits = 3) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

bmi_lo %>%
  pivot_longer(c(lo_pre, lo_dia), names_to = "threshold", values_to = "log_odds") %>%
  mutate(threshold = dplyr::case_match(threshold,
    "lo_pre" ~ "≥ Pre-Diabetes",
    "lo_dia" ~ "≥ Diabetes")) %>%
  ggplot(aes(x = BMI_mid, y = log_odds, colour = threshold)) +
  geom_point(size = 3) + geom_line() +
  labs(x = "BMI (quartile midpoint)", y = "Log-odds P(Y ≥ j)", colour = "Threshold") +
  theme_minimal()
```

**Interpretation:** Pearson correlations between BMI quartile midpoint and empirical log-odds were `r bmi_cor_pre` (≥ Pre-Diabetes) and `r bmi_cor_dia` (≥ Diabetes). `r if(abs(bmi_cor_pre) >= 0.95 & abs(bmi_cor_dia) >= 0.95) { "Both are near-perfect, strongly supporting the log-linear assumption for BMI." } else if(abs(bmi_cor_pre) >= 0.85 & abs(bmi_cor_dia) >= 0.85) { "Both are strong, broadly supporting the assumption. Minor deviation is present but unlikely to meaningfully bias the BMI coefficient." } else { "One or both fall below 0.85. The log-linear assumption for BMI may not hold across the full BMI range, thus the coefficient is interpreted with caution." }`

---

# Model Results

```{r}
#| label: term-labels
#| include: false

term_label <- function(x) {
  dplyr::case_match(x,
    "HighBPYes"           ~ "High Blood Pressure (Yes vs No)",
    "HighCholYes"         ~ "High Cholesterol (Yes vs No)",
    "BMI"                 ~ "BMI (per 1-unit increase)",
    "SmokerYes"           ~ "Ever Smoked (Yes vs No)",
    "PhysActivityYes"     ~ "Physical Activity past 30 days (Yes vs No)",
    "GenHlthVery Good"    ~ "General Health: Very Good vs Excellent",
    "GenHlthGood"         ~ "General Health: Good vs Excellent",
    "GenHlthFair"         ~ "General Health: Fair vs Excellent",
    "GenHlthPoor"         ~ "General Health: Poor vs Excellent",
    "SexMale"             ~ "Sex: Male vs Female",
    "Age35-44"            ~ "Age: 35–44 vs 18–34",
    "Age45-54"            ~ "Age: 45–54 vs 18–34",
    "Age55-64"            ~ "Age: 55–64 vs 18–34",
    "Age65-74"            ~ "Age: 65–74 vs 18–34",
    "Age75+"              ~ "Age: 75+ vs 18–34",
    .default = x
  )
}
```

## Adjusted Odds Ratios

Results are presented as exponentiated coefficients (odds ratios). OR > 1 indicates association with higher odds of more severe diabetes, and OR < 1 is protective. All ORs are adjusted for all other predictors in the model.

```{r}
#| label: or-table

or_tbl <- tidy(m_full, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(coef.type == "location")

n_terms  <- nrow(or_tbl)
n_sig    <- sum(or_tbl$p.value < 0.05, na.rm = TRUE)
nonsig_t <- term_label(or_tbl$term[or_tbl$p.value >= 0.05])

cat(sprintf("Total predictor terms : %d\n", n_terms))
cat(sprintf("Significant (α = 0.05): %d\n", n_sig))
if (length(nonsig_t) > 0)
  cat(sprintf("Non-significant       : %s\n", paste(nonsig_t, collapse = ", ")))

or_tbl %>%
  mutate(term = term_label(term)) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  kable(digits = 3,
        col.names = c("Predictor", "OR", "95% CI Lower", "95% CI Upper", "p-value"),
        caption = "Table 3. Adjusted Odds Ratios : Ordinal Logistic Regression") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(0, bold = TRUE) %>%
  footnote(general = "Reference: HighBP = No, HighChol = No, Smoker = No, PhysActivity = No, GenHlth = Excellent, Sex = Female, Age = 18–34.")
```

**Interpretation:** Of the `r n_terms` predictor terms, `r n_sig` `r ifelse(n_sig == n_terms, "(all)", paste0("of ", n_terms))` are statistically significant at α = 0.05. `r if(length(nonsig_t) > 0) paste0("Non-significant: ", paste(nonsig_t, collapse = ", "), " thus retained on clinical grounds.") else ""` Predictor-level interpretation follows in the Interpretation of Results section.

## Threshold Coefficients

```{r}
#| label: thresholds

thresh_df <- tidy(m_full, conf.int = TRUE) %>% filter(coef.type == "intercept")
t1_val    <- round(thresh_df$estimate[thresh_df$term == "No Diabetes|Pre-Diabetes"], 3)
t2_val    <- round(thresh_df$estimate[thresh_df$term == "Pre-Diabetes|Diabetes"], 3)

cat(sprintf("Threshold 1 (No Diabetes | Pre-Diabetes) = %.3f\n", t1_val))
cat(sprintf("Threshold 2 (Pre-Diabetes | Diabetes)    = %.3f\n", t2_val))

thresh_df %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  kable(digits = 3,
        col.names = c("Threshold", "Estimate", "SE", "z", "p-value"),
        caption = "Table 4. Threshold Coefficients (Cut Points)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**Interpretation:** The first threshold (`No Diabetes|Pre-Diabetes` = `r t1_val`) is the log-odds boundary where cumulative probability shifts from *No Diabetes* into the higher categories. The second (`Pre-Diabetes|Diabetes` = `r t2_val`) marks the boundary into *Diabetes* alone. Since `r t2_val` > `r t1_val`, the thresholds are correctly ordered, thus no reversal, confirming internal model consistency. Both are nuisance parameters and not interpreted clinically.

---

# Proportional Odds Assumption

The **proportional odds assumption** (parallel regression assumption) requires that the OR for each predictor is constant across both cumulative thresholds. Two formal tests are applied using the `ordinal` package.

> **Note on Brant test:** The Brant test (via `MASS::polr()`) is not applied here. With only n = 150 in the Pre-Diabetes category and multi-level predictors (GenHlth: 4 levels, Age: 6 levels), the test consistently fails to converge, thus fitting separate binary logistic models at each threshold becomes computationally infeasible with sparse middle-category cells. `nominal_test()` and `scale_test()` from `ordinal` are the more appropriate assessments for `clm()` models.

## Nominal Test

The nominal test compares the fitted model against one where each predictor is allowed a different effect at each threshold. A significant p-value (p < 0.05) means the effect of that predictor varies across thresholds, which is a violation of proportional odds.

```{r}
#| label: nominal-test

nt_res    <- nominal_test(m_full)
nt_pvals  <- nt_res[["Pr(>Chi)"]]
names(nt_pvals) <- rownames(nt_res)
nt_pvals  <- nt_pvals[names(nt_pvals) != "<none>"]   # exclude baseline row
nt_fail   <- names(nt_pvals[!is.na(nt_pvals) & nt_pvals < 0.05])
nt_pass   <- names(nt_pvals[!is.na(nt_pvals) & nt_pvals >= 0.05])
nt_na     <- names(nt_pvals[is.na(nt_pvals)])

cat(sprintf("Violate assumption (p < 0.05): %s\n",
            if(length(nt_fail) > 0) paste(nt_fail, collapse = ", ") else "none"))
cat(sprintf("Hold assumption   (p ≥ 0.05): %s\n",
            if(length(nt_pass) > 0) paste(nt_pass, collapse = ", ") else "none"))
cat(sprintf("Returned NA                 : %s\n",
            if(length(nt_na) > 0) paste(nt_na, collapse = ", ") else "none"))
nt_res
```

**Interpretation:** `r length(nt_fail)` of `r length(nt_fail) + length(nt_pass)` predictors showed significant deviation from proportional odds (p < 0.05): `r if(length(nt_fail) > 0) paste(nt_fail, collapse = ", ") else "none"`. `r if(length(nt_pass) > 0) paste0(length(nt_pass), " predictor(s) held the assumption: ", paste(nt_pass, collapse = ", "), ".") else ""` Flagged predictors have effects that differ in magnitude across the two thresholds (not necessarily in direction). With n = `r nrow(dat)`, this test is sensitive enough to pick up trivially small deviations, and statistical significance alone does not determine whether the violation is practically meaningful.

## Scale Test

The scale test checks whether the variance of the underlying latent variable differs across predictor levels, indicating heteroscedasticity on the latent scale. A significant result means the spread of the latent diabetes severity distribution is not constant across that predictor's levels.

```{r}
#| label: scale-test

st_res    <- scale_test(m_full)
st_pvals  <- st_res[["Pr(>Chi)"]]
names(st_pvals) <- rownames(st_res)
st_pvals  <- st_pvals[names(st_pvals) != "<none>"]   # exclude baseline row
st_fail   <- names(st_pvals[!is.na(st_pvals) & st_pvals < 0.05])
st_pass   <- names(st_pvals[!is.na(st_pvals) & st_pvals >= 0.05])

cat(sprintf("Significant scale effect (p < 0.05): %s\n",
            if(length(st_fail) > 0) paste(st_fail, collapse = ", ") else "none"))
cat(sprintf("No scale effect          (p ≥ 0.05): %s\n",
            if(length(st_pass) > 0) paste(st_pass, collapse = ", ") else "none"))
st_res
```

**Interpretation:** `r length(st_fail)` predictor(s) showed a significant scale effect (p < 0.05): `r if(length(st_fail) > 0) paste(st_fail, collapse = ", ") else "none"`. `r if(length(st_pass) > 0) paste0(length(st_pass), " predictor(s) showed no scale effect: ", paste(st_pass, collapse = ", "), ".") else ""` `r if(length(st_fail) > 0) "Latent variance is not constant across levels of the flagged predictors, and OR estimates for those predictors carry this caveat." else "No significant scale effects were found. Latent variance appears consistent across predictor levels, supporting the homoscedasticity assumption."`

---

# Model Fit

## Global Likelihood Ratio Test

```{r}
#| label: lrt-global

m_null     <- clm(diabetes_cat ~ 1, data = dat)
lrt_global <- anova(m_null, m_full)
lrt_chi    <- round(lrt_global[2, "LR.stat"], 1)
lrt_df     <- lrt_global[2, "df"]

cat(sprintf("Global LRT: χ²(%d) = %.1f, p < 0.001\n", lrt_df, lrt_chi))
cat(sprintf("AIC — Null: %.0f  |  Full: %.0f  |  ΔAIC: %.0f\n",
            AIC(m_null), AIC(m_full), AIC(m_null) - AIC(m_full)))
lrt_global
```

**Interpretation:** The full model significantly improves fit over the null (LRT χ²(`r lrt_df`) = `r format(lrt_chi, big.mark = ",")`, p < 0.001). AIC dropped from `r round(AIC(m_null), 0)` to `r round(AIC(m_full), 0)`, a reduction of `r round(AIC(m_null) - AIC(m_full), 0)` points. The 8 predictors collectively explain substantial variation in diabetes severity beyond what the intercepts alone capture.

## Pseudo R²

```{r}
#| label: pseudo-r2

loglik_full <- as.numeric(logLik(m_full))
loglik_null <- as.numeric(logLik(m_null))
n           <- nrow(dat)

r2_mcf <- 1 - (loglik_full / loglik_null)
r2_cox <- 1 - exp((2 / n) * (loglik_null - loglik_full))
r2_nag <- r2_cox / (1 - exp((2 / n) * loglik_null))

cat(sprintf("McFadden Pseudo R²   = %.4f\n", r2_mcf))
cat(sprintf("Nagelkerke Pseudo R² = %.4f\n", r2_nag))
```

**Interpretation:** McFadden R² = `r round(r2_mcf, 3)` and Nagelkerke R² = `r round(r2_nag, 3)`. `r if(r2_mcf >= 0.40) { paste0("McFadden R² of ", round(r2_mcf, 3), " exceeds the 0.40 threshold considered excellent for logistic models.") } else if(r2_mcf >= 0.20) { paste0("McFadden R² of ", round(r2_mcf, 3), " falls in the 0.20–0.40 range, indicating good model fit with moderate-to-substantial explained variation.") } else { paste0("McFadden R² of ", round(r2_mcf, 3), " is below 0.20. The current 8 predictors explain a limited proportion of variation in diabetes severity.") }` Nagelkerke R² = `r round(r2_nag, 3)` is the preferred metric in health sciences reporting.

## Variable Contribution (Drop-one LRT)

```{r}
#| label: drop-one

d1_res    <- drop1(m_full, test = "Chi")
d1_pred   <- d1_res[rownames(d1_res) != "<none>", ]
d1_pcol   <- names(d1_pred)[ncol(d1_pred)]
top3      <- paste(term_label(rownames(d1_pred)[order(d1_pred$LRT, decreasing = TRUE)[1:3]]),
                   collapse = ", ")
d1_nonsig <- rownames(d1_pred)[!is.na(d1_pred[[d1_pcol]]) & d1_pred[[d1_pcol]] > 0.05]
n_nonsig  <- length(d1_nonsig)

cat(sprintf("Top 3 contributors: %s\n", top3))
cat(sprintf("Non-significant predictors (p > 0.05): %s\n",
            if(n_nonsig > 0) paste(term_label(d1_nonsig), collapse = ", ") else "none"))
d1_res
```

**Interpretation:** The three largest independent contributions come from **`r top3`**. `r if(n_nonsig == 0) { paste0("All ", nrow(d1_pred), " predictors contributed significantly (p < 0.05) beyond the others and none are redundant in the adjusted model.") } else { paste0(n_nonsig, " predictor(s) did not reach significance in the drop-one test (p > 0.05): ", paste(term_label(d1_nonsig), collapse = ", "), ". Their independent contribution beyond the remaining predictors is marginal in this sample.") }`

---

# Interaction Analysis

One a priori biologically plausible interaction is tested: **`HighBP × BMI`**. Hypertension and obesity share the insulin-resistance pathway, so the per-unit BMI association with diabetes severity may differ between those with and without high blood pressure.

```{r}
#| label: interaction-model

m_int   <- clm(diabetes_cat ~ HighBP * BMI + HighChol + Smoker +
                 PhysActivity + GenHlth + Sex + Age,
               data = dat, link = "logit")

int_lrt <- anova(m_full, m_int)
int_chi <- round(int_lrt[2, "LR.stat"], 2)
int_df  <- int_lrt[2, "df"]
int_p   <- int_lrt[2, "Pr(>Chisq)"]

cat(sprintf("LRT (HighBP × BMI interaction): χ²(%d) = %.2f, p = %.3f\n",
            int_df, int_chi, int_p))
int_lrt
```

```{r}
#| label: interaction-results

int_or_tbl <- tidy(m_int, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(coef.type == "location", str_detect(term, ":"))

int_or_val <- round(int_or_tbl$estimate[1], 3)
int_ci_lo  <- round(int_or_tbl$conf.low[1], 3)
int_ci_hi  <- round(int_or_tbl$conf.high[1], 3)
int_or_p   <- int_or_tbl$p.value[1]

cat(sprintf("HighBP × BMI interaction OR = %.3f (95%% CI: %.3f–%.3f, p = %.3f)\n",
            int_or_val, int_ci_lo, int_ci_hi, int_or_p))

int_or_tbl %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  kable(caption = "Interaction term — OR and 95% CI", digits = 3,
        col.names = c("Term", "OR", "95% CI Lower", "95% CI Upper", "p-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:** Adding the `HighBP × BMI` interaction `r ifelse(int_p < 0.05, "significantly improves", "does not significantly improve")` model fit (χ²(`r int_df`) = `r int_chi`, p = `r round(int_p, 3)`). `r if(int_p < 0.05) { paste0("The interaction OR of ", int_or_val, " (95% CI: ", int_ci_lo, "–", int_ci_hi, ", p = ", round(int_or_p, 3), ") indicates that among those with high blood pressure, each 1-unit BMI increase carries ", int_or_val, " times the per-unit BMI odds compared to those without, indicating a ", ifelse(int_or_val > 1, "stronger", "weaker"), " BMI gradient in the hypertensive group. The interaction model is preferred.") } else { "The main-effects model is retained on parsimony grounds, as the per-unit BMI association does not meaningfully differ by hypertension status in this sample." }`

---

# Model Comparison

```{r}
#| label: model-comparison

m_reduced <- clm(diabetes_cat ~ HighBP + HighChol + BMI + GenHlth + Sex + Age,
                 data = dat, link = "logit")

tibble(
  Model  = c("Null (intercept only)",
             "Reduced (−Smoker, −PhysActivity)",
             "Full model"),
  logLik = c(as.numeric(logLik(m_null)),
             as.numeric(logLik(m_reduced)),
             as.numeric(logLik(m_full))),
  df     = c(attr(logLik(m_null),    "df"),
             attr(logLik(m_reduced), "df"),
             attr(logLik(m_full),    "df")),
  AIC    = c(AIC(m_null), AIC(m_reduced), AIC(m_full)),
  BIC    = c(BIC(m_null), BIC(m_reduced), BIC(m_full))
) %>%
  arrange(AIC) %>%
  kable(digits = 1, caption = "Table 5. Model Comparison — log-likelihood, df, AIC, and BIC") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| label: lrt-reduced

lrt_red      <- anova(m_reduced, m_full)
lrt_red_p    <- lrt_red[2, "Pr(>Chisq)"]
lrt_red_pfmt <- ifelse(lrt_red_p < 0.001, "< 0.001", sprintf("= %.3f", lrt_red_p))
full_best    <- AIC(m_full) < AIC(m_reduced)
delta_aic    <- round(AIC(m_reduced) - AIC(m_full), 1)
delta_bic    <- round(BIC(m_reduced) - BIC(m_full), 1)

cat(sprintf("LRT p-value (Smoker + PhysActivity jointly): p %s\n", lrt_red_pfmt))
lrt_red
```

**Interpretation:** `r if(full_best) { paste0("The full model has the lowest AIC (", round(AIC(m_full),1), ") and BIC (", round(BIC(m_full),1), "). Against the reduced model, ΔAIC = ", delta_aic, " and ΔBIC = ", delta_bic, ".") } else { paste0("The reduced model (AIC = ", round(AIC(m_reduced),1), ", BIC = ", round(BIC(m_reduced),1), ") outperforms the full model (AIC = ", round(AIC(m_full),1), ", BIC = ", round(BIC(m_full),1), ", ΔAIC = ", abs(delta_aic), " favouring reduced). Adding Smoker and PhysActivity does not recover the 2-parameter AIC penalty.") }` The LRT `r if(lrt_red_p < 0.05) { paste0("confirms Smoker and PhysActivity jointly improve fit (p ", lrt_red_pfmt, ") and both are retained.") } else { paste0("shows Smoker and PhysActivity do **not** jointly improve fit beyond the remaining predictors (p ", lrt_red_pfmt, "). The reduced model is statistically preferred. Both are retained here on clinical grounds and their marginal statistical contribution in this sample should be noted as a limitation.") }`

---

# Interpretation of Results

## Proportional Odds Interpretation Framework

In the proportional odds model, the **OR for each predictor is constant across both cut points**:

- The odds of *Pre-Diabetes or Diabetes* vs. *No Diabetes*, **and**
- The odds of *Diabetes* vs. *No Diabetes or Pre-Diabetes*

OR > 1 means greater probability of a more severe category (not a specific threshold).

## Summary Table

```{r}
#| label: or-results

or_results <- tidy(m_full, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(coef.type == "location")

or_results %>%
  mutate(term      = term_label(term),
         sig       = ifelse(p.value < 0.05, "Yes", "No"),
         direction = ifelse(estimate > 1, "Higher risk", "Lower risk")) %>%
  select(term, estimate, conf.low, conf.high, p.value, sig, direction) %>%
  kable(digits = 3,
        col.names = c("Predictor", "OR", "CI Lower", "CI Upper", "p-value",
                      "Significant?", "Direction"),
        caption = "Table 6. Adjusted Odds Ratios — Direction and Significance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| label: or-extract
#| include: false

get_or <- function(trm) {
  x <- filter(or_results, term == trm)
  list(or  = round(x$estimate, 2),
       lo  = round(x$conf.low, 2),
       hi  = round(x$conf.high, 2),
       p   = ifelse(x$p.value < 0.001, "< 0.001", sprintf("= %.3f", x$p.value)),
       dir = ifelse(x$estimate > 1, "higher", "lower"))
}

bp      <- get_or("HighBPYes")
chol    <- get_or("HighCholYes")
bmi     <- get_or("BMI")
smoke   <- get_or("SmokerYes")
pa      <- get_or("PhysActivityYes")
sex     <- get_or("SexMale")
glh_vg  <- get_or("GenHlthVery Good")
glh_g   <- get_or("GenHlthGood")
glh_f   <- get_or("GenHlthFair")
glh_p   <- get_or("GenHlthPoor")
age_old <- get_or("Age75+")
```

> The OR values cited below are drawn from Table 6. All are adjusted for all other predictors in the model.

## Predictor-by-Predictor Interpretation

1. **High Blood Pressure:** Respondents with high blood pressure had **`r bp$or` times higher odds** (95% CI: `r bp$lo`–`r bp$hi`, p `r bp$p`) of more severe diabetes, consistent with the shared insulin-resistance pathway linking hypertension and diabetes.

2. **High Cholesterol:** High cholesterol was associated with **`r chol$or` times the odds** (95% CI: `r chol$lo`–`r chol$hi`, p `r chol$p`) of a higher severity category. Dyslipidaemia is a core component of metabolic syndrome, so this is expected.

3. **BMI:** Each 1-unit increase in BMI multiplied the odds of more severe diabetes by **`r bmi$or`** (95% CI: `r bmi$lo`–`r bmi$hi`, p `r bmi$p`). A 5-unit increase (e.g., BMI 25 → 30) multiplies odds by `r round(bmi$or^5, 2)`, a `r round((bmi$or^5 - 1) * 100, 0)`% increase. BMI is the strongest continuous predictor in the model.

4. **Smoking:** Ever-smoking (≥100 cigarettes lifetime) was associated with **`r smoke$or` times the odds** of more severe diabetes (95% CI: `r smoke$lo`–`r smoke$hi`, p `r smoke$p`). `r if(smoke$or > 1) paste0("An OR of ", smoke$or, ", a ", round((smoke$or - 1)*100, 0), "% increase, is consistent with evidence linking nicotine to beta-cell dysfunction and insulin resistance.") else paste0("An OR of ", smoke$or, " is unexpected (OR < 1). Possible explanations include survivor bias or residual confounding.")`

5. **Physical Activity:** Respondents active in the past 30 days had **`r pa$or` times the odds** of more severe diabetes (95% CI: `r pa$lo`–`r pa$hi`, p `r pa$p`), a **`r round((1 - pa$or) * 100, 0)`% reduction** vs. inactive respondents. This is the largest modifiable behavioural effect in the model.

6. **General Health: Very Good vs Excellent:** OR = **`r glh_vg$or`** (95% CI: `r glh_vg$lo`–`r glh_vg$hi`, p `r glh_vg$p`). A single step down from Excellent to Very Good is linked to `r round(abs(glh_vg$or - 1)*100, 0)`% `r glh_vg$dir` odds.

7. **General Health: Good vs Excellent:** OR = **`r glh_g$or`** (95% CI: `r glh_g$lo`–`r glh_g$hi`, p `r glh_g$p`), `r round(abs(glh_g$or - 1)*100, 0)`% `r glh_g$dir` odds vs. Excellent.

8. **General Health: Fair vs Excellent:** OR = **`r glh_f$or`** (95% CI: `r glh_f$lo`–`r glh_f$hi`, p `r glh_f$p`), `r round(abs(glh_f$or - 1)*100, 0)`% `r glh_f$dir` odds vs. Excellent.

9. **General Health: Poor vs Excellent:** OR = **`r glh_p$or`** (95% CI: `r glh_p$lo`–`r glh_p$hi`, p `r glh_p$p`). The monotone gradient across health levels (Very Good → Good → Fair → Poor: `r glh_vg$or` → `r glh_g$or` → `r glh_f$or` → `r glh_p$or`) is a clear dose-response pattern. Self-rated health appears to work well as a proxy for cumulative metabolic burden here (which is not surprising, as it often does in BRFSS-based analyses).

10. **Sex:** Male respondents had **`r sex$or` times the odds** of more severe diabetes vs. females (95% CI: `r sex$lo`–`r sex$hi`, p `r sex$p`). `r if(sex$or > 1) paste0("An OR of ", sex$or, ", indicating ", round((sex$or - 1)*100, 0), "% higher odds in males, may reflect differences in fat distribution, health-seeking behaviour, or risk factor clustering.") else paste0("An OR of ", sex$or, " indicates ", round((1 - sex$or)*100, 0), "% lower odds in males, possibly reflecting earlier clinical detection due to more frequent metabolic screening.")`

11. **Age 75+ vs 18–34:** The oldest group had **`r age_old$or` times the odds** of more severe diabetes vs. the youngest (95% CI: `r age_old$lo`–`r age_old$hi`, p `r age_old$p`), a `r round((age_old$or - 1)*100, 0)`% increase, reflecting cumulative metabolic ageing, declining beta-cell function, and decades of risk factor exposure. The progressive increase across all age bands in Table 6 confirms a monotone age gradient.

---

# Prediction

## Classification Accuracy

```{r}
#| label: classification-accuracy

pred_class <- predict(m_full, type = "class")$fit
conf_mat   <- table(Predicted = pred_class, Observed = dat$diabetes_cat)
accuracy   <- mean(pred_class == as.character(dat$diabetes_cat))
acc_pct    <- round(accuracy * 100, 1)
per_class  <- round(diag(conf_mat) / colSums(conf_mat) * 100, 1)

cat(sprintf("Overall accuracy        = %.1f%%\n", acc_pct))
cat(sprintf("Sensitivity — No Diabetes   : %.1f%%\n", per_class["No Diabetes"]))
cat(sprintf("Sensitivity — Pre-Diabetes  : %.1f%%\n", per_class["Pre-Diabetes"]))
cat(sprintf("Sensitivity — Diabetes      : %.1f%%\n", per_class["Diabetes"]))
```

```{r}
#| label: confusion-matrix

conf_mat %>%
  as.data.frame() %>%
  pivot_wider(names_from = Observed, values_from = Freq) %>%
  kable(caption = "Table 7. Confusion Matrix — Predicted vs Observed Diabetes Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Interpretation:** The model correctly classified `r acc_pct`% of observations overall. Per-class sensitivity: No Diabetes = `r per_class["No Diabetes"]`%, Pre-Diabetes = `r per_class["Pre-Diabetes"]`%, Diabetes = `r per_class["Diabetes"]`%. `r if(per_class["Pre-Diabetes"] < 30) { paste0("Pre-Diabetes sensitivity is very low (", per_class["Pre-Diabetes"], "%), which is expected given only n = 150 cases flanked by two much larger categories. Most Pre-Diabetes misclassifications fall into adjacent categories.") } else if(per_class["Pre-Diabetes"] < 60) { paste0("Pre-Diabetes sensitivity of ", per_class["Pre-Diabetes"], "% is moderate, reflecting the difficulty of classifying the middle category with n = 150.") } else { paste0("Pre-Diabetes sensitivity of ", per_class["Pre-Diabetes"], "% is reasonable given the small cell size (n = 150).") }`

## Fitted Probabilities for Illustrative Profiles

```{r}
#| label: prediction-profiles

new_profiles <- data.frame(
  HighBP       = factor(c("No",  "Yes", "No",  "Yes"), levels = c("No","Yes")),
  HighChol     = factor(c("No",  "Yes", "No",  "Yes"), levels = c("No","Yes")),
  BMI          = c(22,   32,   22,   32),
  Smoker       = factor(c("No",  "No",  "Yes", "Yes"), levels = c("No","Yes")),
  PhysActivity = factor(c("Yes", "Yes", "No",  "No"),  levels = c("No","Yes")),
  GenHlth      = factor(c("Excellent","Good","Fair","Poor"),
                        levels = c("Excellent","Very Good","Good","Fair","Poor")),
  Sex          = factor(c("Female","Male","Female","Male"), levels = c("Female","Male")),
  Age          = factor(c("35-44","45-54","55-64","65-74"),
                        levels = c("18-34","35-44","45-54","55-64","65-74","75+"))
)

pred_probs <- predict(m_full, newdata = new_profiles, type = "prob")
pp_mat     <- round(pred_probs$fit, 3)

cat(sprintf("Profile 1 — P(No Diabetes) = %.3f, P(Pre-Diabetes) = %.3f, P(Diabetes) = %.3f\n",
            pp_mat[1,1], pp_mat[1,2], pp_mat[1,3]))
cat(sprintf("Profile 4 — P(No Diabetes) = %.3f, P(Pre-Diabetes) = %.3f, P(Diabetes) = %.3f\n",
            pp_mat[4,1], pp_mat[4,2], pp_mat[4,3]))

cbind(data.frame(Profile = paste0("Profile ", 1:4),
                 BMI     = new_profiles$BMI,
                 HighBP  = new_profiles$HighBP,
                 Age     = new_profiles$Age,
                 GenHlth = new_profiles$GenHlth),
      pp_mat) %>%
  kable(col.names = c("Profile", "BMI", "High BP", "Age", "Gen. Health",
                      "P(No Diabetes)", "P(Pre-Diabetes)", "P(Diabetes)"),
        caption = "Table 8. Predicted Probabilities for Four Illustrative Profiles") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:** Profile 1 (female, BMI 22, no metabolic risk, Excellent health, aged 35–44) has P(No Diabetes) = `r pp_mat[1,1]` (`r round(pp_mat[1,1]*100,1)`%). Profile 4 (male, BMI 32, high BP, high cholesterol, smoker, sedentary, Poor health, aged 65–74) has P(Diabetes) = `r pp_mat[4,3]` (`r round(pp_mat[4,3]*100,1)`%). The difference of `r round((pp_mat[4,3] - pp_mat[1,3])*100, 1)` percentage points in P(Diabetes) between these two profiles shows how multiplicative risk accumulates. Probabilities across the three columns sum to 1.0 per row. Note that because Pre-Diabetes was deliberately oversampled (n = 150 vs. approximately 1.8% in the population), these predicted probabilities reflect sample-level associations and will overstate P(Pre-Diabetes) relative to the true population; they should not be applied as population risk estimates.

## Manual Calculation of Predicted Probabilities

Following Hosmer, Lemeshow & Sturdivant (Eq. 8.25), cumulative and category probabilities are derived step-by-step from the model coefficients, then verified against `predict()`.

**Formula:**

$$P(Y \leq j \mid X) = \frac{1}{1 + \exp[-(\alpha_j - X\hat{\beta})]}$$

$$P(Y = j \mid X) = P(Y \leq j) - P(Y \leq j - 1)$$

where $\alpha_j$ is the $j$-th threshold and $X\hat{\beta}$ is the linear predictor summed over all active predictors.

**Illustration: Profile 1** (Female, BMI 22, No HighBP, No HighChol, No Smoker, PhysActivity = Yes, GenHlth = Excellent, Age = 35–44):

Reference categories for binary and categorical predictors are: Female, No HighBP, No HighChol, No Smoker, PhysActivity = No, GenHlth = Excellent, Age = 18–34. For those predictors Profile 1 is at reference. BMI is continuous and has no reference category, so its contribution is BMI × β_BMI = 22 × β_BMI and must be included in the linear predictor.

```{r}
#| label: manual-prediction

coefs  <- coef(m_full)
alpha1 <- coefs["No Diabetes|Pre-Diabetes"]
alpha2 <- coefs["Pre-Diabetes|Diabetes"]

cat(sprintf("Threshold 1 (No Diabetes | Pre-Diabetes)  α₁ = %.4f\n", alpha1))
cat(sprintf("Threshold 2 (Pre-Diabetes | Diabetes)     α₂ = %.4f\n\n", alpha2))

# Active predictors for Profile 1
# Binary/categorical predictors at reference = 0 contribution
# BMI is continuous: contribution = β_BMI × 22
beta_pa  <- coefs["PhysActivityYes"]
beta_age <- coefs["Age35-44"]
beta_bmi <- coefs["BMI"]
xb_p1    <- beta_bmi * 22 + beta_pa * 1 + beta_age * 1

cat(sprintf("β(BMI) × 22           = %.4f × 22 = %.4f\n", beta_bmi, beta_bmi * 22))
cat(sprintf("β(PhysActivity = Yes) = %.4f\n", beta_pa))
cat(sprintf("β(Age 35-44)          = %.4f\n", beta_age))
cat(sprintf("Linear predictor Xβ   = %.4f + %.4f + %.4f = %.4f\n\n",
            beta_bmi * 22, beta_pa, beta_age, xb_p1))

# Step 1: Cumulative probabilities P(Y ≤ j)
p_leq1 <- 1 / (1 + exp(-(alpha1 - xb_p1)))
p_leq2 <- 1 / (1 + exp(-(alpha2 - xb_p1)))

cat(sprintf("P(Y ≤ No Diabetes)   = 1 / (1 + exp(-(%.4f - %.4f))) = %.4f\n",
            alpha1, xb_p1, p_leq1))
cat(sprintf("P(Y ≤ Pre-Diabetes)  = 1 / (1 + exp(-(%.4f - %.4f))) = %.4f\n\n",
            alpha2, xb_p1, p_leq2))

# Step 2: Category probabilities by differencing
p1_no  <- p_leq1
p1_pre <- p_leq2 - p_leq1
p1_dia <- 1 - p_leq2

cat(sprintf("P(No Diabetes)  = P(Y ≤ No Diabetes)                    = %.4f\n", p1_no))
cat(sprintf("P(Pre-Diabetes) = P(Y ≤ Pre-Diabetes) - P(No Diabetes)  = %.4f\n", p1_pre))
cat(sprintf("P(Diabetes)     = 1 - P(Y ≤ Pre-Diabetes)               = %.4f\n", p1_dia))
cat(sprintf("Sum             = %.4f  (must equal 1.000)\n\n", p1_no + p1_pre + p1_dia))

# Verification
cat(sprintf("predict() result: P(No Diabetes) = %.4f\n", pp_mat[1, 1]))
cat(sprintf("Manual result:    P(No Diabetes) = %.4f\n", p1_no))
cat(sprintf("Difference:       %.6f  (rounding only)\n", abs(pp_mat[1, 1] - p1_no)))
```

**Interpretation:** The manual calculation reproduces the `predict()` output to within rounding error, confirming the model applies the proportional odds formula correctly. The linear predictor Xβ for Profile 1 is `r round(coef(m_full)["BMI"] * 22 + coef(m_full)["PhysActivityYes"] + coef(m_full)["Age35-44"], 4)`, composed of the BMI contribution (22 × β_BMI), the PhysActivity = Yes effect, and the Age 35–44 effect, with all other predictors at their reference categories contributing zero. The two threshold parameters then map this single linear predictor onto the three-category probability distribution.

---

# Discussion

## Summary of Findings

Among `r nrow(dat)` US adults from the BRFSS 2015 survey, several independent predictors of diabetes severity emerged after adjustment.

**Metabolic risk factors** including high blood pressure (OR = `r bp$or`), high cholesterol (OR = `r chol$or`), and BMI (OR = `r bmi$or` per unit) carried the strongest associations. The BMI effect compounds quickly: a 5-unit increase multiplies cumulative odds by `r round(bmi$or^5, 2)`, making it the dominant continuous predictor.

Physical activity was associated with a `r round((1 - pa$or) * 100, 0)`% reduction in odds (OR = `r pa$or`), the largest modifiable behavioural effect in the model. That finding is worth emphasising, because it is also the most actionable.

**Self-rated health gradient:** A stepwise increase in odds from Very Good through to Poor (ORs: `r glh_vg$or` → `r glh_g$or` → `r glh_f$or` → `r glh_p$or`) confirmed the expected dose-response. Self-rated health is doing real work here as a proxy for cumulative metabolic load.

Older age groups showed progressively higher odds, reaching `r age_old$or`-fold for the 75+ group. Male sex was associated with `r sex$dir` odds (OR = `r sex$or`).

## Limitations

1. **Cross-sectional design:** BRFSS is cross-sectional and causal inference is not warranted.
2. **Self-reported diabetes status:** Pre-diabetes is frequently undiagnosed, introducing misclassification in the middle category.
3. **Oversampled Pre-Diabetes:** ORs are sample-level associations and population prevalence cannot be derived from this sample.
4. **Proportional odds assumption:** Nominal and scale tests flagged several predictors. The single OR per predictor is a threshold-averaged effect, and a partial proportional odds model would serve as a useful sensitivity analysis.
5. **Residual confounding:** Unmeasured factors (family history, diet quality, medications) may bias estimates.

## Conclusion

BMI, hypertension, dyslipidaemia, and older age were independently associated with higher diabetes severity. Physical activity was protective. The cross-sectional design limits causal claims, but the findings are consistent with integrated metabolic syndrome screening and lifestyle intervention as priorities for at-risk adults.

---

# Software

```{r}
#| label: session-info

sessionInfo()
```

---

# References

1. Hosmer, D.W., Lemeshow, S. & Sturdivant, R.X. (2013). *Applied Logistic Regression*, 3rd ed. Chapter 8. Wiley.
2. Christensen, R.H.B. (2022). `ordinal`: Regression Models for Ordinal Data. R package. https://cran.r-project.org/package=ordinal
3. CDC BRFSS (2015). *Behavioral Risk Factor Surveillance System Survey Data*. U.S. Department of Health and Human Services, CDC, Atlanta, Georgia.
4. UCI Machine Learning Repository. CDC Diabetes Health Indicators Dataset. https://doi.org/10.24432/C53919
5. Brant, R. (1990). Assessing proportionality in the proportional odds model for ordinal logistic regression. *Biometrics*, 46(4), 1171–1178.
