---
title: "Multinomial Logistic Regression Analysis of Contraceptive Method Choice"
subtitle: "1987 National Indonesia Contraceptive Prevalence Survey"
author: "Munir"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    number-sections: true
    theme: flatly
    code-fold: true
    code-summary: "Show Code"
    embed-resources: true
    highlight-style: github
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

Contraceptive method choice in Indonesia is not random. Family planning decisions in the 1980s reflected a mix of education, economics, religion, and age in ways that varied considerably by method type. Understanding which factors drove uptake of long-acting versus short-term methods matters for programme design, looking not just at whether women used contraception, but at what kind.

This analysis uses data from the **1987 National Indonesia Contraceptive Prevalence Survey** (n = 1,473 married non-pregnant women) to examine determinants of method choice. The outcome has three nominal categories: no use, long-term methods (IUD, sterilisation), and short-term methods (pill, condom, injection). No natural clinical ordering exists between the two use categories, so multinomial logistic regression was used with no use as the reference group.

# Methods

## Data Source and Variables

Data were obtained from the 1987 National Indonesia Contraceptive Prevalence Survey, available from the UCI Machine Learning Repository. The analytical sample comprised 1,473 married non-pregnant women.

The **outcome variable** was contraceptive method choice, categorised as no use (reference), long-term method, or short-term method.

**Table 1** describes all predictor variables included in the analysis.

| Variable | Description | Type |
|---|---|---|
| `wife_age` | Wife's age in years | Continuous |
| `wife_edu` | Wife's education: 1=Low, 2=Medium-low, 3=Medium-high, 4=High | Ordinal |
| `husb_edu` | Husband's education: 1=Low, 2=Medium-low, 3=Medium-high, 4=High | Ordinal |
| `num_children` | Number of children ever born | Continuous |
| `wife_religion` | Religion: 0=Non-Islam, 1=Islam | Binary |
| `wife_working` | Working: 0=Yes, 1=No | Binary |
| `husb_occupation` | Husband's occupation: 1=Agriculture, 2=Trade, 3=Service, 4=Other | Nominal |
| `sol_index` | Standard of living index: 1=Low to 4=High | Ordinal |
| `media_exposure` | Media exposure: 0=Good, 1=Not good | Binary |

> Note: `wife_working` carries reversed coding: 0 indicates currently working and 1 indicates not working.

## Statistical Analysis

Multinomial logistic regression was performed using `VGAM::vglm()` with `family = multinomial` in R. No use was set as the reference category. Simple (unadjusted) models were first fitted for wife's religion and wife's education separately. A multivariable model was then fitted including seven predictors: wife's age, wife's education, wife's religion, wife's working status, standard of living index, number of children, and media exposure. Husband's education and husband's occupation were included in Table 1 for descriptive purposes but were not entered into the models, as the analysis focused on the wife's own sociodemographic and household characteristics. Wife's education and the standard of living index were each dichotomised into low (levels 1–2) versus high (levels 3–4) for the final model, after inspecting the coefficient pattern in the full four-level version.

Results are presented as Relative Risk Ratios (RRR) with 95% confidence intervals (CI) and p-values. A p-value < 0.05 was considered statistically significant. Predicted probabilities were derived for specific covariate profiles using the multinomial probability formula. Findings were cross-validated using `nnet::multinom()`. All analyses were conducted in R.

### Data Preparation

```{r}
#| label: libraries

library(here)
library(tidyverse)
library(gtsummary)
library(VGAM)
library(nnet)
library(knitr)
library(kableExtra)
```

```{r}
#| label: read-data

cmc <- read_csv("cmc.csv")
```

```{r}
#| label: glimpse

glimpse(cmc)
summary(cmc)
```

```{r}
#| label: recode-factors

cmc <- cmc %>%
  mutate(
    cmc_method     = factor(contraceptive_method,
                            labels = c("no_use", "long_term", "short_term")),
    wife_religion2 = factor(wife_religion,
                            labels = c("non_islam", "islam")),
    wife_working2  = factor(wife_working,
                            labels = c("yes", "no")),
    media2         = factor(media_exposure,
                            labels = c("good", "not_good")),
    wife_edu2      = factor(wife_edu,
                            labels = c("low", "med_low", "med_high", "high")),
    husb_edu2      = factor(husb_edu,
                            labels = c("low", "med_low", "med_high", "high")),
    sol_index2     = factor(sol_index,
                            labels = c("low", "med_low", "med_high", "high")),
    husb_occ2      = factor(husb_occupation,
                            labels = c("Agriculture", "Trade", "Service", "Other"))
  )
```

```{r}
#| label: recode-dichotomous

cmc <- cmc %>%
  mutate(
    wife_edu_d  = factor(ifelse(wife_edu  <= 2, "low", "high"), levels = c("low", "high")),
    sol_index_d = factor(ifelse(sol_index <= 2, "low", "high"), levels = c("low", "high"))
  )

cmc %>% count(wife_edu, wife_edu_d)
cmc %>% count(sol_index, sol_index_d)
```

```{r}
#| label: set-reference

cmc <- cmc %>%
  mutate(
    # VGAM requires no_use as the LAST level
    cmc_vgam = fct_relevel(cmc_method, c("long_term", "short_term", "no_use")),
    # nnet requires no_use as the FIRST level
    cmc_nnet = relevel(cmc_method, ref = "no_use")
  )

levels(cmc$cmc_vgam)
levels(cmc$cmc_nnet)
```

# Results

## Sample Characteristics

```{r}
#| label: tbl-descriptive
#| tbl-cap: "Table 2. Sample characteristics by contraceptive method"

cmc %>%
  mutate(
    cmc_method     = factor(contraceptive_method,
                            labels = c("No use", "Long-term", "Short-term")),
    wife_religion2 = factor(wife_religion, labels = c("Non-Islam", "Islam")),
    wife_working2  = factor(wife_working,  labels = c("Yes", "No")),
    media2         = factor(media_exposure, labels = c("Good", "Not good")),
    wife_edu2      = factor(wife_edu,  labels = c("Low", "Med-low", "Med-high", "High")),
    husb_edu2      = factor(husb_edu,  labels = c("Low", "Med-low", "Med-high", "High")),
    sol_index2     = factor(sol_index, labels = c("Low", "Med-low", "Med-high", "High")),
    husb_occ2      = factor(husb_occupation,
                            labels = c("Agriculture", "Trade", "Service", "Other"))
  ) %>%
  select(wife_age, wife_edu2, husb_edu2, num_children,
         wife_religion2, wife_working2, husb_occ2,
         sol_index2, media2, cmc_method) %>%
  tbl_summary(
    by = cmc_method,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      wife_age       ~ "Wife's age (years)",
      wife_edu2      ~ "Wife's education",
      husb_edu2      ~ "Husband's education",
      num_children   ~ "No. of children ever born",
      wife_religion2 ~ "Religion",
      wife_working2  ~ "Wife currently working",
      husb_occ2      ~ "Husband's occupation",
      sol_index2     ~ "Standard of living index",
      media2         ~ "Media exposure"
    ),
    missing = "no"
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  italicize_levels()
```

Of the 1,473 women, **629 (42.7%)** used no contraception, **333 (22.6%)** used a long-term method, and **511 (34.7%)** used a short-term method. Long-term users were older on average and had more children, while short-term users had the lowest mean age. Educational attainment was visibly skewed toward the higher categories in the long-term group. Islamic religion was more common in the no-use group than among long-term users (refer to Table 2 for group-level distributions).

## Unadjusted Analysis

### Simple Model 1: Wife's Religion

> A simple (unadjusted) multinomial logistic regression with wife's religion as the sole predictor establishes the unadjusted association between religion and method choice before controlling for anything else.

```{r}
#| label: model1-fit

fitmlog1 <- vglm(cmc_vgam ~ wife_religion2,
                 multinomial,
                 data = cmc)

summary(fitmlog1)
```

```{r}
#| label: model1-rrr

b_m1  <- coef(fitmlog1)
ci_m1 <- confint(fitmlog1)

b_rrr_m1    <- cbind(b_m1, exp(b_m1))
ci_b_rrr_m1 <- cbind(ci_m1, exp(ci_m1))
tab_m1      <- cbind(b_rrr_m1, ci_b_rrr_m1)

colnames(tab_m1) <- c("b", "rrr",
                       "lower 95% b", "upper 95% b",
                       "lower 95% rrr", "upper 95% rrr")

kable(round(tab_m1, 3),
      caption = "Table 3. Unadjusted model, Wife's Religion: Log Odds, RRR and 95% CI") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(3:4, background = "#f0f7ff")
```

**Long-term vs No-use:** Islamic women had substantially lower relative odds of long-term use than non-Islamic women (RRR = 0.46, 95% CI: 0.32–0.65, p < 0.001).

**Short-term vs No-use:** No significant difference between religious groups (RRR = 0.87, 95% CI: 0.61–1.23, p = 0.425).

Religion suppressed long-term use strongly at the unadjusted level, with no parallel effect on short-term use.

### Simple Model 2: Wife's Education

> Wife's education is entered as a four-level categorical predictor to check for a dose-response pattern, specifically whether higher education is progressively associated with greater contraceptive use relative to no use.

```{r}
#| label: model2-fit

fitmlog2 <- vglm(cmc_vgam ~ wife_edu2,
                 multinomial,
                 data = cmc)

summary(fitmlog2)
```

```{r}
#| label: model2-rrr

b_m2  <- coef(fitmlog2)
ci_m2 <- confint(fitmlog2)

b_rrr_m2    <- cbind(b_m2, exp(b_m2))
ci_b_rrr_m2 <- cbind(ci_m2, exp(ci_m2))
tab_m2      <- cbind(b_rrr_m2, ci_b_rrr_m2)

colnames(tab_m2) <- c("b", "rrr",
                       "lower 95% b", "upper 95% b",
                       "lower 95% rrr", "upper 95% rrr")

kable(round(tab_m2, 3),
      caption = "Table 4. Unadjusted model, Wife's Education: Log Odds, RRR and 95% CI") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

**Long-term vs No-use** (reference: low education):

- Medium-low education: RRR = 2.41 (p = 0.025)
- Medium-high education: RRR = 5.23 (p < 0.001)
- High education: RRR = 13.54 (p < 0.001)

**Short-term vs No-use** (reference: low education):

- Medium-low education: RRR = 1.77 (p = 0.010)
- Medium-high education: RRR = 2.28 (p < 0.001)
- High education: RRR = 2.87 (p < 0.001)

A clear dose-response gradient is present for both method types. The RRRs for long-term use are far larger (13.54 at the top level versus 2.87 for short-term), which suggests education specifically drives uptake of long-acting methods, not just contraceptive use in general.

## Adjusted Analysis

### Full Model

> All seven predictors are entered simultaneously to obtain adjusted estimates. Four-level factors are kept at this stage to inspect the coefficient pattern before deciding whether to simplify.

```{r}
#| label: model3-fit

fitmlog3 <- vglm(
  cmc_vgam ~ wife_age + wife_edu2 + wife_religion2 +
             wife_working2 + sol_index2 + num_children + media2,
  multinomial,
  data = cmc
)

summary(fitmlog3)
```

Coefficients for wife's education and standard of living showed broadly monotone effects across the four levels, supporting dichotomisation (low = levels 1–2, high = levels 3–4) for the final model.

### Final Model

> Wife's education and standard of living index are each collapsed to two levels (low vs high) to improve parsimony while preserving the main contrast. This is the model used for final inference and predictions.

```{r}
#| label: model4-fit

fitmlog4 <- vglm(
  cmc_vgam ~ wife_age + wife_edu_d + wife_religion2 +
             wife_working2 + sol_index_d + num_children + media2,
  multinomial,
  data = cmc
)

summary(fitmlog4)
```

### Model Fit

```{r}
#| label: model-fit

m_null  <- vglm(cmc_vgam ~ 1, multinomial, data = cmc)
ll_full <- as.numeric(logLik(fitmlog4))
ll_null <- as.numeric(logLik(m_null))
df_full <- attr(logLik(fitmlog4), "df")
df_null <- attr(logLik(m_null),   "df")

r2_mcf  <- 1 - ll_full / ll_null
lrt_chi <- -2 * (ll_null - ll_full)
lrt_df  <- df_full - df_null
lrt_p   <- pchisq(lrt_chi, df = lrt_df, lower.tail = FALSE)

cat(sprintf("Log-likelihood (null model) = %.2f\n", ll_null))
cat(sprintf("Log-likelihood (full model) = %.2f\n", ll_full))
cat(sprintf("McFadden R²                 = %.4f\n", r2_mcf))
cat(sprintf("LRT chi-sq(%d)              = %.2f\n", lrt_df, lrt_chi))
cat(sprintf("LRT p-value                 = %.6f\n", lrt_p))
```

The full model significantly improved fit over the null (intercept-only) model (LRT $\chi^2$(`r lrt_df`) = `r round(lrt_chi, 2)`, p < 0.001). McFadden R² = `r round(r2_mcf, 3)`. In multinomial logistic regression, McFadden values of 0.20–0.40 are considered good fit and values above 0.40 are considered excellent; values below 0.20 indicate that the predictors explain a limited share of the variation in method choice.

### Estimation

> Log-odds coefficients (`b`) and their exponentiated forms (Relative Risk Ratios, `rrr`) are extracted, combined with 95% confidence intervals, and displayed in a single table. RRR > 1 indicates higher relative odds of that method over no use, and RRR < 1 indicates lower relative odds.

```{r}
#| label: tbl-final-results

b_m4  <- coef(fitmlog4)
ci_m4 <- confint(fitmlog4)

b_rrr_m4    <- cbind(b_m4, exp(b_m4))
ci_b_rrr_m4 <- cbind(ci_m4, exp(ci_m4))
res_m4      <- cbind(b_rrr_m4, ci_b_rrr_m4)

colnames(res_m4) <- c("b", "rrr",
                       "lower 95% b", "upper 95% b",
                       "lower 95% rrr", "upper 95% rrr")

kable(round(res_m4, 3),
      caption = "Table 5. Adjusted multivariable model: Log Odds, RRR and 95% CI") %>%
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed", "responsive")) %>%
  pack_rows("Long-term vs No-use", 1, nrow(res_m4) / 2) %>%
  pack_rows("Short-term vs No-use", nrow(res_m4) / 2 + 1, nrow(res_m4))
```

### Inference

> The table above produces two sets of estimates, one per comparison equation. Each predictor's adjusted effect from Table 5 is interpreted below.

**Long-term vs No-use** (adjusted)

- **Wife's age**: RRR = 0.97 (95% CI: 0.95–0.99, p = 0.003), indicating a 3% decrease in relative odds of long-term use per additional year of age.
- **Wife's education** (high vs low): RRR = 4.69 (95% CI: 3.20–6.90, p < 0.001), the strongest predictor, with highly educated women having nearly five times the relative odds.
- **Wife's religion** (Islam vs non-Islam): RRR = 0.56 (95% CI: 0.38–0.82, p = 0.003), indicating Islamic women had 44% lower relative odds of long-term use.
- **Wife's working status** (no vs yes): RRR = 0.97 (95% CI: 0.74–1.44, p = 0.848), not statistically significant.
- **Standard of living** (high vs low): RRR = 2.37 (95% CI: 1.57–3.57, p < 0.001), indicating women with a higher standard of living had more than double the relative odds.
- **Number of children**: RRR = 1.35 (95% CI: 1.25–1.46, p < 0.001), with each additional child associated with a 35% increase in relative odds.
- **Media exposure** (not good vs good): RRR = 0.50 (95% CI: 0.24–1.05, p = 0.068), directionally suggestive but not significant.

**Short-term vs No-use** (adjusted)

- **Wife's age**: RRR = 0.90 (95% CI: 0.88–0.92, p < 0.001), a stronger inverse effect than for long-term, as younger women preferentially chose reversible methods.
- **Wife's education** (high vs low): RRR = 1.78 (95% CI: 1.35–2.36, p < 0.001), significant but markedly smaller than the long-term effect (RRR 4.69 vs 1.78).
- **Wife's religion** (Islam vs non-Islam): RRR = 0.65 (95% CI: 0.44–0.96, p = 0.028), indicating Islamic women had 35% lower relative odds of short-term use.
- **Wife's working status** (no vs yes): RRR = 1.14 (p = 0.380), not statistically significant.
- **Standard of living** (high vs low): RRR = 1.52 (95% CI: 1.12–2.04, p = 0.006), a significant positive association.
- **Number of children**: RRR = 1.40 (95% CI: 1.30–1.51, p < 0.001), with each additional child associated with a 40% increase in relative odds.
- **Media exposure** (not good vs good): RRR = 0.52 (95% CI: 0.31–0.88, p = 0.015), significant, with poor media access acting as a barrier to short-term use and no comparable effect for long-term use.

## Prediction

> Prediction is demonstrated using the simple Model 1 (wife's religion only) to keep calculations transparent. `type = "link"` returns predicted log-odds and `type = "response"` converts these to probabilities on the 0–1 scale.

### Predicted Log-odds

```{r}
#| label: predict-logodds

head(predict.vgam(fitmlog1, type = "link"))
```

- Column 1 = predicted log-odds of **long-term vs no-use** and Column 2 = predicted log-odds of **short-term vs no-use**.
- Only two unique rows appear (one for non-Islamic women and one for Islamic women) because this model has a single binary predictor.
- Islamic women show a more negative value in Column 1, consistent with the negative religion coefficient (b = −0.7813).

### Predicted Probabilities

```{r}
#| label: predict-prob

head(predict.vgam(fitmlog1, type = "response"))
```

- Each row contains three predicted probabilities (no-use, long-term, short-term) that sum to exactly **1.0**.
- Two distinct profiles appear, corresponding to non-Islamic and Islamic women.
- Islamic women show a higher predicted probability of no-use and a lower probability of long-term use, directly reflecting RRR = 0.46 from the model above.

### Predicted Probabilities for Specific Profiles

> The multinomial probability formula is applied manually to verify the model output and show how predicted probabilities are computed from log-odds coefficients.

Manual probability estimation was performed for two hypothetical women, one non-Islamic and one Islamic, using the multinomial probability formula:

$$P(\text{category } j) = \frac{\exp(\eta_j)}{1 + \sum_{k=1}^{J-1} \exp(\eta_k)}$$

where $\eta_j$ denotes the linear predictor for category $j$ and $\eta_J = 0$ for the reference category (no-use).

```{r}
#| label: manual-prediction

int1   <- coef(fitmlog1)["(Intercept):1"]
int2   <- coef(fitmlog1)["(Intercept):2"]
b_rel1 <- coef(fitmlog1)["wife_religion2islam:1"]
b_rel2 <- coef(fitmlog1)["wife_religion2islam:2"]

# Non-Islamic woman (wife_religion = 0)
logit1_nonislam <- int1 + b_rel1 * 0
logit2_nonislam <- int2 + b_rel2 * 0

denom_nonislam <- 1 + exp(logit1_nonislam) + exp(logit2_nonislam)

p_noUse_ni <- 1                    / denom_nonislam
p_long_ni  <- exp(logit1_nonislam) / denom_nonislam
p_short_ni <- exp(logit2_nonislam) / denom_nonislam

cat("── Non-Islamic woman ──────────────────────────────\n")
cat("  Log-odds Long-term vs No-use  :", round(logit1_nonislam, 4), "\n")
cat("  Log-odds Short-term vs No-use :", round(logit2_nonislam, 4), "\n")
cat("  P(No use)     =", round(p_noUse_ni, 4), "\n")
cat("  P(Long-term)  =", round(p_long_ni,  4), "\n")
cat("  P(Short-term) =", round(p_short_ni, 4), "\n")
cat("  Sum           =", round(p_noUse_ni + p_long_ni + p_short_ni, 4), "\n\n")

# Islamic woman (wife_religion = 1)
logit1_islam <- int1 + b_rel1 * 1
logit2_islam <- int2 + b_rel2 * 1

denom_islam <- 1 + exp(logit1_islam) + exp(logit2_islam)

p_noUse_is <- 1                 / denom_islam
p_long_is  <- exp(logit1_islam) / denom_islam
p_short_is <- exp(logit2_islam) / denom_islam

cat("── Islamic woman ──────────────────────────────────\n")
cat("  Log-odds Long-term vs No-use  :", round(logit1_islam, 4), "\n")
cat("  Log-odds Short-term vs No-use :", round(logit2_islam, 4), "\n")
cat("  P(No use)     =", round(p_noUse_is, 4), "\n")
cat("  P(Long-term)  =", round(p_long_is,  4), "\n")
cat("  P(Short-term) =", round(p_short_is, 4), "\n")
cat("  Sum           =", round(p_noUse_is + p_long_is + p_short_is, 4), "\n")
```

**Non-Islamic woman** (`wife_religion = 0`):

- Log-odds (long-term vs no-use) = 0.0132, log-odds (short-term vs no-use) = −0.0834
- P(no-use) = 0.3409, P(long-term) = 0.3455, P(short-term) = 0.3136
- Probabilities are roughly equal across all three categories, with a sum of 1.0

**Islamic woman** (`wife_religion = 1`):

- Log-odds (long-term vs no-use) = 0.0132 + (−0.7813) = −0.7681, log-odds (short-term vs no-use) = −0.0834 + (−0.1425) = −0.2259
- P(no-use) = 0.4421, P(long-term) = 0.2051, P(short-term) = 0.3528
- Islamic women have a substantially higher predicted probability of no-use and lower long-term use, with a sum of 1.0

Both vectors match `predict.vgam(fitmlog1, type = "response")`, confirming the manual calculation is correct.

## Cross-validation with nnet::multinom

Results were cross-validated using `nnet::multinom()`. As `nnet` does not report p-values in its summary, p-values were derived from the ratio of each coefficient to its standard error under a normal approximation. Confidence intervals are returned as a three-dimensional array, with one slice per outcome category.

```{r}
#| label: nnet-simple

mlog_nnet <- multinom(cmc_nnet ~ wife_religion2,
                      data  = cmc,
                      trace = FALSE)

summary(mlog_nnet)
```

```{r}
#| label: nnet-pvalues

z_stat <- summary(mlog_nnet)$coefficients /
          summary(mlog_nnet)$standard.errors

p_val <- (1 - pnorm(abs(z_stat), 0, 1)) * 2

kable(round(p_val, 4),
      caption = "Table 6. P-values from nnet::multinom, z-test") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
#| label: nnet-ci

ci_nnet <- confint(mlog_nnet, level = 0.95)

cat("=== Long-term vs No-use ===\n")
print(round(ci_nnet[,, "long_term"], 3))

cat("\n=== Short-term vs No-use ===\n")
print(round(ci_nnet[,, "short_term"], 3))
```

```{r}
#| label: compare-packages

mlog_nnet_full <- multinom(
  cmc_nnet ~ wife_age + wife_edu_d + wife_religion2 +
             wife_working2 + sol_index_d + num_children + media2,
  data  = cmc,
  trace = FALSE
)

cat("=== Coefficients: VGAM::vglm (fitmlog4) ===\n")
print(round(coef(fitmlog4), 4))

cat("\n=== Coefficients: nnet::multinom (mlog_nnet_full) ===\n")
print(round(coef(mlog_nnet_full), 4))
```

Coefficient estimates from `nnet::multinom` were numerically identical to those from `VGAM::vglm`. Both packages maximise the same multinomial log-likelihood, and any minor floating-point differences reflect different numerical optimisation routines, not substantive analytical differences.

# Discussion

Among 1,473 married non-pregnant Indonesian women in 1987, education was the dominant predictor of long-term contraceptive use. Highly educated women had nearly five times the relative odds of long-term use compared to women with low education (RRR = 4.69), against an effect of only 1.78 for short-term use. That gap is not trivial, suggesting education does something specific for long-acting method acceptance, not just contraceptive uptake broadly.

**Number of children** showed similar-magnitude effects for both method types (long-term RRR = 1.35, short-term RRR = 1.40 per child). Parity appears to drive contraceptive adoption in general, without much method-specific preference.

**Standard of living** was positive for both comparisons but stronger for long-term use (RRR = 2.37 vs 1.52). Access and cost probably explain part of this, though the cross-sectional data cannot separate financial capacity from other correlated factors.

**Religion** reduced relative odds for both method types, more so for long-term use (RRR = 0.56) than for short-term (RRR = 0.65). In the 1980s Indonesian Islamic context, this pattern is consistent with accepted birth spacing but resistance to permanent or long-acting methods (which is worth noting as a historically specific finding, not a generalisation about contemporary practice).

**Wife's age** showed a stronger inverse effect for short-term use (RRR = 0.90 per year) than long-term (RRR = 0.97), suggesting younger women leaned toward reversible options.

**Media exposure** was significant only for short-term use (RRR = 0.52, p = 0.015), not for long-term (p = 0.068). Mass media channels appear more relevant to short-term method uptake, though the confidence interval for long-term use is wide and that null finding is not definitive.

**Wife's working status** was not significant for either comparison after adjustment.

# Limitations

Several limitations should be noted. First, the data are cross-sectional, so causal inferences cannot be drawn from these findings. Second, contraceptive method was self-reported, which may introduce recall or social desirability bias, particularly for methods that carry religious or cultural sensitivity. Third, the data are from 1987 and specific to Indonesia; findings may not generalise to other countries or to contemporary Indonesian populations where education levels, family planning programmes, and religious practice have changed considerably. Fourth, husband's education and occupation were available but were not included in the final model; residual confounding from unmeasured partner-level characteristics therefore cannot be excluded.

# References

1. Musa KI. *Multinomial Logistic Regression: Course Notes for Advanced Categorical Statistics*. Universiti Sains Malaysia, 2022.

2. Hosmer DW, Lemeshow S, Sturdivant RX. *Applied Logistic Regression*, 3rd ed. Wiley, 2013.

3. UCI Machine Learning Repository. *Contraceptive Method Choice Data Set*. 1987 National Indonesia Contraceptive Prevalence Survey. <https://archive.ics.uci.edu/ml/datasets/Contraceptive+Method+Choice>

4. Long JS, Freese J. *Regression Models for Categorical Dependent Variables Using Stata*, 3rd ed. Stata Press, 2014.

5. Yee TW. *Vector Generalized Linear and Additive Models: With an Implementation in R*. Springer, 2015.
